---
title: "Chapter 2: Working with Date and Time Objects"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

# The date and time formats

# Date and time objects in R

The `base` package provides two types of date and time classes:

1.  `Date`: this basic format puts dates in the `YYYY-m-d` format and tracks the number of days since the default of `1970-01-01`.
2.  `POSIXct`/`POSIXlt`: Known as a `DateTime` class and they represent the calendar date, the time of day, and the time zone using the ISO 8601 international standard format of `YYYY-m-d H:M:S`. `POSIXct` represents the number of seconds since the origin point as a numeric vector and `POSIXlt` stores each one of the date and time elements as a list.

It makes sense to use a time object only if the series frequency is higher than daily. Otherwise, it is recommended that you use the date object. Let's review the difference between these objects:

```{r}
date <- Sys.Date()
date

time_ct <- Sys.time()
time_ct
```

Check out the class of each object:

```{r}
class(date)

class(time_ct)
```

We can convert the `POSIXct` object to a `POSIXlt` object with `as.POSIClt`:

```{r}
time_lt <- as.POSIXlt(time_ct)
time_lt
class(time_lt)
```

Both of these object appear to be the same, but the difference is in how they are stored internally:

```{r}
unclass(time_ct)

unclass(time_lt)
```

## Creating date and time objects

We can convert the `"2014-5-12"` into a`Date` object like this:

```{r}
date <- as.Date("2014-5-12")
date
```

Or as a `as.POSIXct` or `as.POSIXlt` objects:

```{r}
time_ct <- as.POSIXct("2014-5-12 20:05:35", tz = "EST")
time_ct
```

## Importing date and time objects

## Reformatting and converting date objects

```{r}
library(tidyverse)

dates_df <- read_csv("https://raw.githubusercontent.com/PacktPublishing/Hands-On-Time-Series-Analysis-with-R/master/Chapter02/dates_formats.csv")

str(dates_df)
```

Let's work to convert each of these variables to the correct date format.

The first variable is the Japanese format for date. Because this is similar to the ISO 8601 standard we can just use `as.Date()`:

```{r}
dates_df$Japanese_format_new <- as.Date(dates_df$Japanese_format)

select(dates_df, Japanese_format_new, Japanese_format)
```

Here is a summary table for the main arguments for date objects:

| Symbol | Meaning                                                         | Example                  |
|--------|-----------------------------------------------------------------|--------------------------|
| `%a`   | Abbreviated weekday name in the current locale on this platform | Sun, Mon, Thu            |
| `%A`   | Full weekday name in the current locale                         | Sunday, Monday, Thursday |
| `%b`   | Abbreviated month name in the current locale on this platform   | Jan, Feb, Mar            |
| `%B`   | Full month name in the current locale                           | January, February, March |
| `%d`   | Day of the month as a decimal number                            | 01, 02, 03               |
| `%m`   | Month as a decimal numner                                       | 01, 02, 03               |
| `%y`   | A year without a century (two-digit)                            | 18                       |
| `%Y`   | A year with a century (four-digit)                              | 2018                     |
| `%r`   | For a 12-hour clock defined by the AM/PM indicator              | AM                       |

Let's format the next variable, the US date format:

```{r}
dates_df$US_format_new <- as.Date(dates_df$US_format, format = "%m/%d/%Y")
select(dates_df, US_format, US_format_new)
```

Now the rest of the columns:

```{r}
dates_df <- 
  dates_df %>% 
  mutate(US_long_format_new = as.Date(US_long_format, format = "%A, %B %d, %Y"),
         CA_mix_format_new = as.Date(CA_mix_format, format = "%B %d, %Y"),
         SA_mix_format_new = as.Date(SA_mix_format, format = "%d %B %Y"),
         NZ_format_new = as.Date(NZ_format, format = "%d/%m/%Y"))

glimpse(dates_df)
```

## Handling numeric date objects

Date objects in R are stored as numeric values from an origin point (01-01-1970 as the default in R). This means that we can correctly identify dates in this format if we know the right origin point. For Excel, it's 12/30/1899.

```{r}
dates_df$Excel_Numeric_Format_new <- as.Date(dates_df$Excel_Numeric_Format, origin = as.Date("1899-12-30"))

select(dates_df, Excel_Numeric_Format_new, Excel_Numeric_Format)
```

Let's check our work:

```{r}
str(dates_df)
```

## Reformatting and conversion of time objects

The `POSIX` classes are an extension of the `Date` class, with the addition of four elements: hours, minutes, seconds and time zone. Let's start with an example using the ISO 8601 international standard:

```{r}
time_str <- "2018-12-31 23:59:59"

time_posix_ct1 <- as.POSIXct(time_str)
time_posix_ct1
```

The numeric value of a time object represents the number of seconds since the origin point.

```{r}
time_numeric <- 1546318799

time_posix_ct2 <- as.POSIXct(time_numeric, origin = "1970-01-01")
time_posix_ct2
```

In addition to mapping the date and time, we also have to identify and map the cycle of the day (AM or PM).

```{r}
time_US_str <- "Monday, December 31, 2018 11:59:59 PM"
```

Gets formatted by:

```{r}
time_posix_ct3 <- as.POSIXct(time_US_str, format = "%A, %B %d, %Y %I:%M:%S %p")
time_posix_ct3
```

Here is a table for formatting time objects:

| Symbol | Meaning                                                            | Example       |
|--------|--------------------------------------------------------------------|---------------|
| `%H`   | Hours as a decimal number (00-23)                                  | 17:00:00      |
| `%I`   | Hours as a decimal number (01-12)                                  | 5:00:00 PM    |
| `%M`   | Minutes as a decimal number (00-59)                                | 17:15:00      |
| `%S`   | Seconds as an integer (00-61), allowing for up to two leap-seconds | 17:15:45      |
| `%p`   | AM/PM indicator in the locale                                      | AM, PM        |
| `%r`   | Equivalent to `%I:%M:%S %p` for 12-hour clock time                 | 9:35:00 PM    |
| `%T`   | Equivalent to `%H:%M:%S`                                           | 17:15:00      |
| `%Z`   | Time zone abbreviation as a character string                       | GMT, EST, PDT |

## Time zone setting

Time zone is the 7th element of the `POSIX` classes and be set with the `tz` argument of the `format` argument.

# Creating a date or time index

The `base` package provides two functions for producing time index vectors with `Date` or `POSIX` objects respectively `seq.Date()` and `seq.POSIXt()`. `seq.Date()` is good for daily frequencies or lower, `seq.POSIXt()` is good for higher frequencies like hourly, etc.

```{r}
daily_index <- seq.Date(from = as.Date("2016-01-01"),
                        to = as.Date("2018-12-31"),
                        by = "day")

head(daily_index)
```

# Manipulation of date and time with the lubridate package

## Reformatting date and time objects - the lubridate way

Let's go back to the complex object:

```{r}
time_US_str <- "Monday, December 31, 2018 11:59:59 PM"
```

To convert this with lubridate, we use the `ymd_hms()` function:

```{r}
library(lubridate)

time_lubridate <- mdy_hms(time_US_str, tz = "EST")
time_lubridate
```

Let's go back to the original data and convert the first 6 variables into date objects with the `ymd()` function:

```{r}
dates_df <- read_csv("https://raw.githubusercontent.com/PacktPublishing/Hands-On-Time-Series-Analysis-with-R/master/Chapter02/dates_formats.csv")

dates_df <- 
  dates_df %>% 
  mutate(Japanese_format_new = ymd(Japanese_format),
         across(US_format:CA_mix_format, ~ mdy(.x), .names = "{.col}_new"),
         across(SA_mix_format:NZ_format, ~ dmy(.x), .names = "{.col}_new"))
```

To convert numeric values to date objects, we use `as_date()` for date objects and `as_datetime()` for time objects.

```{r}
dates_df$Excel_Numeric_Format_new <- as_date(dates_df$Excel_Numeric_Format,
                                             origin = ymd("1899-12-30"))
```

## Utility functions for date and time objects

lubridate provides tools to extract elements of time and date options:

```{r}
time_obj <- mdy_hms(time_US_str, tz = "EST")

# day of the year
yday(time_obj)

# day of the quarter
qday(time_obj)

# day of the month
mday(time_obj)

# day of the week
wday(time_obj, label = TRUE)
```
