---
title: "Chapter 2: The Time Series Object"
output: html_notebook
---

R provides several classes for representing time series objects for a variety of applications. Among those classes, `ts` is one of the main formats for time series data in R, mainly due to its simplicity and the wide adoption of this class by the main packages in R for time series analysis, for example, the `forecast` and `stats` packages.

# The Natural Gas Consumption dataset

```{r}
library(pacman)
p_load(Quandl)

ngc <- Quandl(code = "FRED/NATURALGAS",
              collapse = "quarterly",
              type = "ts",
              end_date = "2018-12-31")

class(ngc)
```

The simplest method to plot a `ts` object is with the `plot` function:

```{r}
plot.ts(ngc,
        main = "US Quarterly Natural Gas Consumption",
        ylab = "Billion of Cubic Feet")
```

# The attributes of the `ts` class

A regular time series is defined as an ordered sequence of observations over time, which is captured at equally spaced time intervals. Whenever this condition ceases to exist, the series becomes an irregular time series. The main characteristics of regular time series data is as follows:

-   Cycle/period: a regular unit of time that split the series into consecutive and equally long subsets

-   frequency: defines the length or the number of units of the cycle

-   timestamp: provides the time each observation in the series was captured, and can be used as the series index.

A `ts` object is composed of two elements - the series values and its corresponding timestamp.

```{r}
# number of observaitions
length(ngc)
```

We can look at the structure of a `ts` dataset with the `head()` function:

```{r}
ngc
```

Here the rows represent the number of the cycle and the columns represent the cycle units. For the `ngc` data, each calendar year is a full cycle and the quarters are the cycle units.

The `cycle()` and the `time()` functions from the **stats** package provide the cycle units and the timestamp of each observation in the series:

```{r}
cycle(ngc)

time(ngc)
```

A more concise way to get this information is with the `frequency()` and `deltat()` functions:

```{r}
frequency(ngc)

deltat(ngc)
```

Other useful functions are `start()` and `end()`:

```{r}
start(ngc)

end(ngc)
```

The `ts_info()` function from the **TStudio** package provides a concise summary of most of the functions above.

```{r}
p_load(TSstudio)

ts_info(ngc)
```

## Multivariate time series objects

When you have multivariate time series data, you need to use the `mts` (multiple time series) class. This combines the functionality of the `ts` and `matrix` classes.

```{r}
data("Coffee_Prices")
head(Coffee_Prices)

ts_info(Coffee_Prices)
```

## Creating a `ts` object

```{r}
my_ts1 <- ts(data = 1:60,
             start = c(2010, 1),
             end = c(2014, 12),
             frequency = 12)

ts_info(my_ts1)

my_ts1
```

Now we will work through the typical process of converting data from a `data.frame` to a `ts` object.

```{r}
library(tidyverse)

# load the data
data("US_indicators")
str(US_indicators)
```

For now, we will only convert the vehicle sales into a `ts` object.

```{r}
tvs <- 
  US_indicators %>% 
  select(Date, `Vehicle Sales`) %>% 
  arrange(Date)

head(tvs)
```

Next, we need to define the start or end of the series. In this case, the series started in January 1976 so we can define it as `start = c(1976, 1)`. Or we can write code to capture the starting point.

```{r}
library(lubridate)

start_point <- c(year(min(tvs$Date)), month(min(tvs$Date)))
start_point
```

Now we build the series:

```{r}
tvs_ts <- ts(data = tvs$`Vehicle Sales`,
             start = start_point,
             frequency = 12)
```

One of the main limitations of the `ts` class is that it can only support two input elements for the timestamp. For example, when we converted `tvs` into a `ts` object, we lost the day component because `ts` could only store the month and year.

## Creating an `mts` object

```{r}
US_indicators <- arrange(US_indicators, Date)

US_indicators_ts <- ts(data = select(US_indicators, `Vehicle Sales`, 
                                     `Unemployment Rate`),
                       start = c(year(min(tvs$Date)), month(min(tvs$Date))),
                       frequency = 12)

ts_info(US_indicators_ts)
```

## Setting the series frequency

Setting the frequency of a series sets the length of a cycle.

$$
\text{Frequency} = \frac{\text{cycle length}}{\text{time interval between observation}}
$$

In this example we will see how setting the frequency impacts the structure of the `ts` object output. First, we simulate close to ten years of daily data.

```{r}
daily_df <- data.frame(date = seq.Date(from = as.Date("2010-01-01"),
                                       length.out = 365 * 10, by = "day"),
                       y = rnorm(365 * 10, mean = 15, sd = 2))

str(daily_df)
```

Create `ts` object:

```{r}
days_week_ts <- ts(daily_df,
                   start = c(1, wday(min(daily_df$date))),
                   frequency = 7)

ts_info(days_week_ts)
```

# Data manipulation of `ts` objects

## The window function

The main purpose of a window function is to subset a `ts` object based on a time range. The main argument of the `window()` function are the `start` and `end` arguments. Let's use the `window()` function to extract all the observations of the year 2005 from the NGC series:

```{r}
window(ngc, start = c(2005, 1), end = c(2005, 4))
```

We can also extract a specific frequency unit from the series. Say we're interested in extracting all the observations of the series that occurred in the third quarter of the year. This can be done by setting the starting point at the third quarter of the first year and the `frequency` to 1.

```{r}
window(ngc, start = c(2000, 3), frequency = 1)
```

## Aggregating `ts` objects

The `aggregate()` function splits the data into subsets, computes specific summary statistics, and then aggregates the results to a `ts` or `data.frame` object. Let's use `aggregate()` to transform the NGC series from a quarterly frequency to yearly:

```{r}
ngc_yearly <- aggregate(ngc, nfrequency = 1, FUN = "sum")
ngc_yearly
```

## Creating lags and leads for `ts` objects

The `lag()` function from the **stats** package (this should not be confused with the `lag()` function from the **dplyr** package) can be used to create lags or leads for `ts` objects.

```{r}
ngc_lag4 <- stats::lag(ngc, k = -4)

ts_info(ngc_lag4)
```

# Visualizing `ts` and `mts` objects

## The `plot.ts()` function

Plotting a `ts` object:

```{r}
plot.ts(tvs_ts,
        main = "US Monthly Total Vehicle Sales",
        ylab = "Thousands of Vehicle",
        xlab = "Time")
```

Plotting an `mts` object:

```{r}
plot.ts(US_indicators_ts,
        plot.type = "multiple",
        main = "US Monthly Vehicle Sales vs. Unemployment Rate",
        xlab = "Time")
```

## The **dygraphs** package

The **dygraphs** package is an R interface to the `dygraphs` JavaScript charting library.

```{r}
p_load(dygraphs)

dygraph(tvs_ts,
        main = "US Monthly Total Vehicle Sales",
        ylab = "Thousands of Vehicle") %>% 
  dyRangeSelector()
```

For the `US_indicators_ts` series, we will add a second *y*-axis, which allows us to plot and compare the two series that are not on the same scale:

```{r}
dygraph(US_indicators_ts,
        main = "US Monthly Vehicle Sales vs. Unemployment Rate") %>% 
  dyAxis("y", label = "Vehicle Sales") %>% 
  dyAxis("y2", label = "Unemployment Rate") %>% 
  dySeries("Vehicle Sales", axis = "y", color = "green") %>% 
  dySeries("Unemployment Rate", axis = "y2", color = "red") %>% 
  dyLegend(width = 400)
```

## The TSstudio package

```{r}
p_load(TSstudio)

ts_plot(tvs_ts,
        title = "US Monthly Total Vehicle Sales",
        Ytitle = "Thousands of Vehicle",
        slider = TRUE)
```

We can add an interactive slider for the *x*-axis.

```{r}
ts_plot(US_indicators_ts,
        title = "US Monthly Vehicle Sales vs. Unemployment Rate",
        type = "multiple")
```
