---
title: "Chapter 3: Working with `zoo` and `xts` Objects"
output: html_notebook
---

# The `zoo` class

The **zoo** package provides a framework for working with regular and irregular time series data. It is comprised of two components:

-   Data structure: a vector or matrix format

-   Index vector: this stores the series observation's corresponding index

Unlike the `ts` class, the index of the `zoo` class can store different date and time classes, such as `Date`, `POSIXct/lt`, `yearmon`, or `yearqtr`, as indices.

```{r}
library(TSstudio)
library(zoo)

data("EURO_Brent")

ts_info(EURO_Brent)
```

## The `zoo` class attributes

Pretty much the same as with `ts` objects.

## The index of the `zoo` object

```{r}
head(index(EURO_Brent))
```

`zoo` objects have their own `attributes()` and `class()` functions:

```{r}
class(index(EURO_Brent))

attributes(index(EURO_Brent))
```

## Working with date and time objects

The **zoo** package provides a set of utility functions for handling and preprocessing date and time objects. The package also comes with two types of `date` objects, `yearmon` and `yearqtr`, for indexing a monthly and quarterly series, respectively.

Let's create a sequence of 12 dates, representing the beginning of each month in 2019, with the `seq.Date` function:

```{r}
monthly_dates <- seq.Date(from = as.Date("2019-01-01"), length.out = 12, 
                          by = "month")

head(monthly_dates)
```

Next, we use the `as.yearmon()` function to convert the `monthly_dates` object to a `yearmon` object:

```{r}
monthly_yearmon <- as.yearmon(monthly_dates)
head(monthly_yearmon)
```

The `as.Date.ts()` function allows us to convert the time stamp of a `ts` object to a `Date` object when applicable.

```{r}
data(USgas)

head(time(USgas))
```

Now let's look at the results of the `as.Date.ts()` function:

```{r}
head(as.Date.ts(USgas))
```

## Creating a `zoo` object

```{r}
data(US_indicators)
str(US_indicators)

vehicle_sales1 <- zoo(x = US_indicators$`Vehicle Sales`,
                      frequency = 12)
head(vehicle_sales1)
```

The `order.by` argument is used to set the `index` operator of the output object.

```{r}
vehicle_sales2 <- zoo(x = US_indicators$`Vehicle Sales`,
                      order.by = US_indicators$Date,
                      frequency = 12)
head(vehicle_sales2)
```

We can also use `as.zoo()` to convert series to `zoo` objects:

```{r}
data(USgas)

usgas_zoo <- as.zoo(USgas)
ts_info(usgas_zoo)
```

## Working with multiple time series objects

We create a multivariate time series object by combining multiple columns:

```{r}
us_indicators_zoo <- zoo(x = US_indicators[ , c("Vehicle Sales", "Unemployment Rate")],
                         frequency = 12,
                         order.by = US_indicators$Date)

ts_info(us_indicators_zoo)

head(us_indicators_zoo)
```

# The xts class

the **xts** package provides the `xts` class and a set of functions and tools for preprocessing, manipulating, and visualizing time series data.

```{r}
data("Michigan_CS")

ts_info(Michigan_CS)
```

## The xts class attributes

## The xts functionality

#### The periodicity function

```{r}
library(xts)

periodicity(Michigan_CS)
```

#### Manipulating the object index

```{r}
tclass(Michigan_CS)
```

We can convert eh object index class with `coverIndex`.

```{r}
Michigan_CS <- convertIndex(Michigan_CS, "Date")
tclass(Michigan_CS)
```

#### Subsetting an xts object based on the index properties

## Manipulating the zoo and xts objects

### Merging time series objects

To merge a series, we need to make sure that the indices have the same structure.

```{r}
tclass(Michigan_CS)

class(index(EURO_Brent))

index(EURO_Brent) <- as.Date(index(EURO_Brent))
class(index(EURO_Brent))
```

Now we need to check for overlap between the indices of the two objects.

```{r}
ts_info(Michigan_CS)

ts_info(EURO_Brent)
```

Now we merge.

```{r}
xts_merge_outer <- merge.xts(Michigan_CS = Michigan_CS,
                             EURO_Brent = EURO_Brent,
                             join = "outer")

ts_info(xts_merge_outer)
```

### Rolling Windows

The `rollapply()` function allows you to apply a function over a specified window of data.

```{r}
EURO_Brent_3m <- rollapply(EURO_Brent, width = 3, FUN = mean)
ts_info(EURO_Brent_3m)
```

### Creating Lags

### Aggregating the zoo and xts objects

We can change the granularity of the data with `aggregate`.

```{r}
usgas_zoo_qtr <- aggregate(usgas_zoo, by = as.yearqtr, FUN = sum)

library(lubridate)
usgas_zoo_year <- aggregate(usgas_zoo, by = year, FUN = sum)
```

## Plotting zoo and xts objects

### The plot.zoo function

```{r}
plot.zoo(EURO_Brent, 
         main = "Crude Oil Prices: Brent - Europe",
         ylab = "USD per Barrel",
         col = "blue")
```

### The plot.xts function

```{r}
plot.xts(Michigan_CS)
```

Now let's creat a summary plot of the `USgas` data. The plot will include:

-   The series

-   a moving average of the series

-   the monthly percentage change of the series

-   the year-over-year percentage growth of each observation of the series

```{r}
usgas_xts <- as.xts(USgas)
tclass(usgas_xts)

# calculate moving average
usgas_xts_ma <- rollapply(usgas_xts, width = 12, FUN = mean)

# merge the two series
usgas_merge <- merge.xts(usgas = usgas_xts,
                         usgas_smooth = usgas_xts_ma)

# calculate the monthly percentage difference and the year-over-year growth
usgas_monthly_diff <- 100 * (usgas_xts / lag(usgas_xts, n = 1) - 1)
usgas_yoy_diff <- 100 * (usgas_xts / lag(usgas_xts, n = 12) - 1)

# plot
plot.xts(usgas_merge,
         main = "US Natural Gas Consumption Summary",
         multi.panel = FALSE,
         col = c("black", "blue"),
         ylim = c(1400, 3700))
lines(usgas_monthly_diff,
      col = "red",
      type = "h",
      on = NA,
      main = "Monthly Difference (%)")
lines(usgas_yoy_diff,
      col = "purple",
      type = "h",
      on = NA,
      main = "YoY Growth (%)")
```

```{r}
library(tidyverse)
ggplot(Michigan_CS, aes(x = Index, y = data)) +
  geom_line() +
  scale_x_date(date_labels = "%b-%Y", date_breaks = "4 year")
```
