---
title: "Chapter 6: Correlation Analysis"
output: html_notebook
---

# Correlation between two variables

# Lags analysis

The goal of lags analysis is to identify and quantify the relationship between a series and its lags. When a series has a seasonal pattern, there should be a correlation with its seasonal lags.

```{r}
library(TSstudio)
data(USgas)

ts_plot(USgas)
```

This series has strong seasonal patterns, but the `EURO_Brent` does not:

```{r}
data("EURO_Brent")

ts_plot(EURO_Brent)
```

The last series, `USVSales` has strong seasonal and cycle patterns:

```{r}
data("USVSales")

ts_plot(USVSales)
```

# The autocorrelation function

This quantifies the relationship between a series and its lags.

```{r}
acf(USgas, lag.max = 60)
```

The blue dotted lines indicate whether the level of correlation between the series and each lag is significant or not. This shows a strong positive correlation with the seasonal lags along with negative correlation with the mid-seasonal lags.

```{r}
acf(EURO_Brent, lag.max = 60)
```

This acf shows a cyclical pattern in the data.

```{r}
acf(USVSales, lag.max = 60)
```

This shows both a seasonal and cycle pattern in the data.

# The partial autocorrelation function

One of the downsides of the acf is that it doesn't remove the effect of lags 1 up to k-1 on the series when calculating the correlation on the series with the k lag. The pacf computes the conditional correlation of the series with the k lag given the relationship of the 1, 2, ..., and k-1 lags. It shows the correlation of the series with the $k$ lag after removing the correlation of the $k$ lag with previous lags.

```{r}
pacf(USgas, lag.max = 60)
```

# Lag plots

Lag plots visualize the relationshup between a series and its lags.

```{r}
ts_lags(USgas)
```

# Causality analysis

The main pitfall of using only past lags is that the lags do not necessarily drive the changes in the series - there could be other exogenous factors. The goal of causality analysis is to identify whether a causal relationship exists between the series we wish to forecast and other potential exogenous factors.

## Causality versus correlation

Two variables will have a causal relationship when the change in one variable triggers a direct change of the second variable.

## The cross-correlation function

The ccf measures the level of correlation between to series and their lags in a fairly similar way as the acf.

```{r}
data("USUnRate")

ts_plot(USUnRate)
```

Let's align the `USUnRate` and `USVSales` series with a `window` function:

```{r}
us_vsales <- window(USVSales, start = c(1976, 1), end = c(2018, 6))
us_unrate <- window(USUnRate, start = c(1976, 1), end = c(2018, 6))

library(plotly)

plot_ly(x = time(us_vsales),
        y = us_vsales) %>% 
  add_lines(x = time(us_unrate),
            y = us_unrate,
            yaxis = "y2") %>% 
  layout(yaxis2 = list(overlaying = "y", side = "right"))
```

We can measure the level of correlation between the unemployment rate and the vehicle sales and its lags using the `ccf()` function.

```{r}
ccf(x = us_vsales, y = us_unrate, lag.max = 36)
```

You can also use `ccf_plot()` from the **TSstudio** package.

```{r}
ccf_plot(x = USVSales, y = USUnRate, lags = 0:12)
```
